# Missing values

```{r, warning=FALSE,message=FALSE}
library(tidyverse)
library(readxl)
library(readr)
library(stringr)
library(ggplot2)
library(GGally)
library(mi)
```

## Inequality Missing Data


```{r  warning = FALSE,  message = FALSE} 
inequality_df <- read_csv("./Inequality/inequality_tidy.csv")
```

Below is a bar plot showing the count of missing data per feature.

```{r fig.width = 8}
prop_missing_features <- data.frame(var = names(colMeans(is.na(inequality_df))), 
            prop_missing = unname(colMeans(is.na(inequality_df)))) %>%
          filter(prop_missing > 0) %>%
          filter(!var %in% c('Country', 'HDI Rank (2018)'))

prop_missing_features %>%
    ggplot(aes(x = fct_reorder(var, prop_missing, .desc =FALSE), y = prop_missing)) +
    geom_col() +
    coord_flip() +
    xlab("") +
    ylab("Proportion of Data Missing") +
    ggtitle("Proportion of Data Missing for Each Inequality Feature")
```


The above shows that for 7 of the 9 features, 20-25% of its entries are missing.

Since there is only 9 years and 9 features, a heat map is an appropriate to visualize combinations of missing data between `Year` and the different features to see if there's any relationship in proportion of missing data for a specific feature across years (and vice versa). 

```{r fig.height = 6, fig.width = 8}

prop_missing_year<- inequality_df %>% 
  group_by(Year) %>% 
  summarise_all(~ sum(is.na(.))/n()) %>%
  ungroup() %>% 
  select(-Country, -'HDI Rank (2018)') %>% 
  pivot_longer(-Year, names_to = 'features' , values_to = 'year_prop_missing') 

prop_missing_year%>%
  ggplot(aes(x = factor(Year), y = fct_reorder(features, -year_prop_missing, mean))) +
  geom_tile(aes(fill = year_prop_missing)) +
  scale_fill_continuous(high = "#132B43", low = "#56B1F7") +
  xlab("Year") +
  ylab('') +
  ggtitle('Heat Map of Missing Data for Inequality \nFeature - Year Combination') +
  labs(fill='Proportion Missing') 
  
```


The above heat map is drawn such that the feature with the largest proportion of missing data is at the bottom. From the heat map, it looks like 2011 and 2012 have the highest rates of missing data for every features aside from "Inequality in life expectancy (%)" and "Inequality-adjusted life expectancy index". However, note that the proportion of missing data peaks at around 0.3 and most feature-year combinations have proportion of missing data that are below 0.2, which are manageable.  

The below is the aggregate sum of number of missing entries over all features for each country, filtering only the countries with at least one missing entry.  

```{r fig.height = 12, fig.width = 8}

prop_missing_countries<- inequality_df %>% 
  group_by(Country) %>% 
  summarise_all(~ sum(is.na(.))/n()) %>%
  ungroup()

prop_missing_countries <- data.frame(Country = prop_missing_countries$Country, 
                                     country_prop_missing =  rowMeans(prop_missing_countries[,4:12])) %>% 
                          filter(country_prop_missing > 0) 

prop_missing_countries %>% 
    ggplot(aes(x = fct_reorder(Country, country_prop_missing, .desc =FALSE), y = country_prop_missing)) +
    geom_col() +
    coord_flip() + 
    xlab('') +
    ylab('Proportion Missing') +
    ggtitle('Proportion of Missing Inequality Data for Each Country')
    
```


100% of the data is missing for Saint Kitts and Nevis, Liechtenstein, and Dominica. In addition, >87.5% of the data is missing for Nauru, Marshall Islands, Andorra, and Tuvalu. All of these countries are extremely small (most are islands) and it is possible that `Inequality` data is not collected nationally and/or data collection is not prioritized by the relevant international organizations.  


Due to the large volume of countries, a heat map is not an effective way to visualize the relationship in proportion of data missing between countries and features. A parallel coordinate plot was drawn instead to determine if there's any relationship between features and proportion of missing data for different countries.

```{r, fig.height = 8, fig.width=10}

inequality_df[,c(1,2,3,4,12,6, 7,10, 5, 9, 11, 8)] %>% 
  group_by(Country) %>% 
  summarise_all(~ sum(is.na(.))/n()) %>%
  ungroup() %>% 
  select(-Year, -'HDI Rank (2018)', -Country) %>%
  rename("Coefficient\nof Human\nInequality" = "Coefficient of Human Inequality",
        "Overall loss\n in HDI\n due to \ninequality\n(%)" = "Overall loss in HDI due to inequality (%)",
        "Inequality\nadjusted \nHDI\n(IHDI)" = "Inequality-adjusted HDI (IHDI)",
        "Inequality\nadjusted\nincome\nindex" = "Inequality-adjusted income index",
        "Inequality\nin\nincome\n(%)" = "Inequality in income (%)",
        "Inequality\nadjusted\neducation\nindex" = "Inequality-adjusted education index",
        "Inequality in \neducation\n(%)" = "Inequality in education (%)",
        "Inequality\nin life\nexpectancy\n(%)" = "Inequality in life expectancy (%)",
        "Inequality\nadjusted\nlife\nexpectancy\nindex" = "Inequality-adjusted life expectancy index"
         ) %>%
  ggparcoord(alpha = 0.2,
             scale = 'globalminmax',
             splineFactor = 0)  +
  theme(axis.text.x = element_text(angle = 0)) +
  xlab("") +
  ylab("") + 
  ggtitle("Inequality Proportion of Missing Data Parallel Coordinate Plot") +
        theme(plot.title = element_text(size = 18),
              axis.text.x = element_text(size = 12)) 

```


* Most countries appear to not be missing data for "Inequality-adjusted life expectancy index" and "Inequality in life expectancy (%)" and when they do, both features are missing in similar proportions. 

* For remaining features, at intervals of around 0.1 from 0 to 1, there are clusters of countries missing that proportion of data for each feature. 

* There is a small number of countries missing data for most to all of the features (flat line at the top), these are likely the smaller countries found in the bar charts.

* For "Coefficient of Human Inequality", "Overall loss in HDI due to inequality (%)", and "Inequality-adjusted HDI (IHDI)", countries generally miss the same proportion of the data for all 3 features, indicating that they may be related. 

* For "Inequality-adjusted income index" and "Inequality in income (%)", countries generally miss the same proportion of the data for these two features; they are likely related. In addition, countries generally are missing an equal to or smaller proportion of data for these two features than the 3 features mentioned in the previous bullet. 

* For "Inequality-adjusted education index" and "Inequality in education (%)" countries generally miss the same proportion of the data for these two features; they are likely related.



## Human Security Missing Data

A significant proportion of data in the Human.Security dataframe is missing.  Largely, this seems to be a product of uneven tracking between variables and geographic complications.

```{r, echo=F}
Human.Security <- read_csv("./HumanSecurity/HumanSecurity.Final.Tidy.csv")
Human.Security <- subset(Human.Security, select = -c(3))
Human.Security <- Human.Security[rowSums(is.na(Human.Security[,3:6])) != ncol(Human.Security[,3:6]),]
Homicide.rate <- read_csv("./HumanSecurity/Homicide.tidy.csv")
Homicide.rate <- Homicide.rate[,2:14]
Homicide.rate.1 <- read.csv("./HumanSecurity/Homicide.tidy.csv")
Homicide.rate.1 <- Homicide.rate.1[,2:14]
names(Homicide.rate.1)[2:13] <- c("1990", "1995", "2000", "2005", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017")
Suicide.rate.F <- read_csv("./HumanSecurity/SR.F.tidy.csv")
Suicide.rate.F <- Suicide.rate.F[,2:6]
Suicide.rate.M <- read_csv("./HumanSecurity/SR.M.tidy.csv")
Suicide.rate.M <- Suicide.rate.M[,2:6]
Prison.popn <- read.csv("./HumanSecurity/Prison.population.csv")
Prison.popn <- subset(Prison.popn, select = -c(X, X.1, X.2, X.3, X.4, X.5, X.6, X.7, X.8, X.9))
names(Prison.popn)[2:11] <- c("2003", "2005", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017")
dcol <- 2
for (i in 2:ncol(Prison.popn)) {
  Prison.popn[,dcol] <- as.numeric(Prison.popn[,dcol])
  dcol <- dcol +1
}
```

The 'Human Security' dataset shows uneven tracking between variables. 'Homicide rate' is recorded in 1990, 1995, 2000, 2005, and yearly from 2010 to 2017. 'Suicide rate' for both males and females, by contrast, were recorded only in the years 2000, 2010, 2015, and 2016.

Below are the years in the 'Homicide rate' dataset for which data were collected:

```{r}
colnames(Homicide.rate)[2:13]
```

By contrast, 'Suicide rate' data for both males and females were only collected for the following years:

```{r}
colnames(Suicide.rate.F)[2:5]
```

The difference in frequency of data collection between these datasets results in a merged dataframe with a high proportion of missing values. 

Specifically, the consistency of the 'Prison population' and 'Homicide rate' data introduces years into the dataframe for which 'Suicide rate' shows no recorded data. As a result, 'Suicide.rate.M' and 'Suicide.rate.F' each show missing values over more than 60% of the dataframe, as shown in the graph below.

```{r}
prop_missing_features <- data.frame(var = names(colMeans(is.na(Human.Security))), 
            prop_missing = unname(colMeans(is.na(Human.Security)))) %>%
          filter(prop_missing > 0)  

prop_missing_features %>%
    ggplot(aes(x = fct_reorder(var, prop_missing, .desc =FALSE), y = prop_missing)) +
    geom_col() +
    coord_flip() +
    xlab(NULL) + 
    ylab("Proportion of Data Missing") +
    ggtitle("Missing Data", subtitle = "Including all years") + 
    ylim(c(0,.9))
```

Alternately, the graph below gives a more accurate picture of the proportion of missing data for each variable.


```{r, echo=F}
hs_missing <- as.data.frame(matrix(nrow = 4, ncol = 2))
colnames(hs_missing) <- c("factor", "prop.missing")
hs_missing[,1] <- c("Homicide.rate", "Prison.popn", "Suicide.F", "Suicide.M")
hs_missing[1,2] <- mean(as.matrix(is.na(Homicide.rate)))
hs_missing[2,2] <- mean(as.matrix(is.na(Prison.popn)))
hs_missing[3,2] <- mean(as.matrix(is.na(Suicide.rate.F)))
hs_missing[4,2] <- mean(as.matrix(is.na(Suicide.rate.M)))

hs_missing %>%
    ggplot(aes(x = fct_reorder(factor, prop.missing, .desc =FALSE), y = prop.missing)) +
    geom_col() +
    coord_flip() +
    xlab(NULL) + 
    ylab("Proportion of Data Missing") +
    ggtitle("Missing Data", subtitle = "Including only collected years for each variable") +
    ylim(c(0,.9))
```

This chart reflects the average proportion of missing data for each variable (across all recorded years and countries). Calculating these proportions separately excludes the empty values introduced by the unequal frequency of data collection between variables. The variables with the greatest proportion of missing data (Prison population and Homicide rate) show between 25% and 35% missing data, respectively.

The following heatmap shows missing data split by variable.  Importantly, the map reflects the entire Human Security dataset, thus again highlighting trends in missing data due to uneven frequency of data collection.

```{r}
prop_missing_year<- Human.Security %>% 
  group_by(Year) %>% 
  summarise_all(~ sum(is.na(.))/n()) %>%
  ungroup() %>% 
  select(-Country) %>% 
  pivot_longer(-Year, names_to = 'features' , values_to = 'year_prop_missing') 

prop_missing_year%>%
  ggplot(aes(x = factor(Year), y = fct_reorder(features, -year_prop_missing, mean))) +
  geom_tile(aes(fill = year_prop_missing)) +
  scale_x_discrete(expand=c(0,0)) +
  scale_fill_continuous(high = "#132B43", low = "#56B1F7") +
  xlab("Year") + 
  ylab(NULL) +
  ggtitle("Missing Human Security Data by Variable")
```

Clearly, 'Suicide.rate.M' and 'Suicide.rate.F' are missing the greatest amount of data, though there seems to be relatively little missing data in the reported years.  Both 'Homicide.rate' and 'Prison.popn' are consistently reported between 2005 and 2017, with slight variations in missing data.  'Homicide.rate' shows no data reported for the year 2003, while 'Prison.popn' shows no data before 2003.

A number of unique problems may be incurred in the collection of international data, as discussed briefly in the Data chapter. Certain countries or regions may have presented challenges for data collection, given that collection processes relied on state institutions and infrastructure.

The 'Homicide.rate' dataset is a particularly demonstrative example.  The final entries in the dataset show countries for which there is no recorded homicide data.

These countries are:

```{r}
print(Homicide.rate$Country[168:192])
```

Of these 24 countries, 20 (Cóte d'Ivoire, Congo, Djibouti, Ethiopia, Gabon, Gambia, Madagascar, Mali, Nigeria, Sudan, Chad, Congo [Dem. Republic], Comoros, Eritrea, Guinea, Equatorial Guinea, Libya, Mauritania, Senegal, and Togo) are in Africa.  The remaining four (Lao [People's Dem. Republic], Vanuatu, Micronesia [Fed. States], the Marshall Islands, and Palau) are small island countries in Oceania.  Potential association between geographical region and missing data represents an important direction for further research, and provides an example of possible trends in missing data.


```{r fig.height = 18, fig.width = 8}

prop_missing_countries <- Human.Security %>% 
  group_by(Country) %>% 
  summarise_all(~ sum(is.na(.))/n()) %>%
  ungroup()

prop_missing_countries <- data.frame(Country = prop_missing_countries$Country, 
                                     country_prop_missing =  rowMeans(prop_missing_countries[,3:6])) %>% 
                          filter(country_prop_missing > 0)

prop_missing_countries %>% 
    ggplot(aes(x = fct_reorder(Country, country_prop_missing, .desc =FALSE), y = country_prop_missing)) +
    geom_col() +
    coord_flip() + 
    xlab(NULL) + 
    ylim(c(0,0.9)) +
    ylab("Proportion of Data Missing") +
    ggtitle("Missing Human Security Data by Country")
```

The above graph shows the proportion of missing data among the Human Security variables by country. 

Note that San Marino, Palau, Nauru, and the Marshall Islands are each missing 75% of data.

Due to the large volume of countries included in this data, and their alphabetical organizational scheme, a heatmap is not an appropriate visualization tool to examine patterns of missing data by country.  


## Education Missing Data


```{r warning = FALSE,  message = FALSE}

df_ed_td<-read_csv('./Education/Education_Tidy.csv')
df_ed_lg<-read_csv('./Education/Education_LG.csv')
df<-df_ed_td[c(2,4:28)] %>% group_by(Country) %>% select(everything()) %>%
     summarise_all(funs(sum(is.na(.)))) 
df$TotalMiss<-rowSums(df[,2:26])
df$PerTot<-df$TotalMiss*100/sum(df$TotalMiss)

```


There are `r sum(df$TotalMiss)` missing observations across the variables, year and country. The following bar chart displays an interesting yet confusing pattern. It looks like all countries have between `r min(df$TotalMiss)` and  `r max(df$TotalMiss)`, corresponding to `r min(df$PerTot)` and `r max(df$PerTot)` percent of total missings. Keeping in mind the data spans 29 years, and 25 variables, this is sensible, though there doesn't seem to be much of a pttern to missing observations. Some smaller countries do seem partially more likely to have missing data.

```{r,fig.height=25,fig.width=10}
ggplot(df,aes(y=fct_reorder(Country,PerTot),x=PerTot)) +
geom_col()+ylab("Country")+xlab("")+ggtitle("Total Missing") +
  theme(plot.title = element_text(hjust = 0.5))
```


For this analysis we want the variables with the least missing values across Countries to have a complete analysis. The chart is too large, and messy to display,but some insights can be gleaned from looking it over that the read is free to explore. Overall it looks like, we can already begin narrowing down some of our variables to the ones that have almost no missings across all countries such as education index, or mean years of schooling. Other interesting notes are that the former Yugoslav Republic of Macedonia seems to have all missing observations-- which is correct according to the original source data, and after a little digging is because it has been replaced by North Macedonia, and is likely a left over data point. Percent of students without access to internet seems to be reported by very few, and mostly developed countries. The percent of population with at least some secondary education also seems like a variable that is frequently reported.

```{r,fig.height=25,fig.width=20,eval=FALSE}

ggplot(df[,1:26] %>%
     pivot_longer(cols=!c("Country"),names_to="Vari",values_to="Val") ,aes(y=Country,x=Vari,fill=Val))+
  geom_tile()+xlab("Variable")+ggtitle("Missing by variable and Country")+theme(plot.title = element_text(hjust = 0.5))+
  scale_fill_continuous(high = "#132B43", low = "#56B1F7")

```


Breaking down the analysis a little further, it looks we should definitely focus on 2010 onwards given the lack of data across countries on a consistent basis for singular years beforehand. It looks like the data was previously released in 5-year buckets. The chart is included to display the yearly pattern across country, though it is still a bit messy.

```{r,fig.height=25,fig.width=20}

df<-df_ed_lg %>% pivot_wider(names_from="Year",values_from="HDI")
df_n<-df[c(2,4:32)] %>% group_by(Country) %>% select(everything()) %>%
     summarise_all(funs(sum(is.na(.)))) 
ggplot(df_n %>%
     pivot_longer(cols=!c("Country"),names_to="Year",values_to="Val") ,aes(y=Country,x=Year,fill=Val))+
  geom_tile()+xlab("Year")+ggtitle("Missing by Year")+theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_continuous(high = "#132B43", low = "#56B1F7")


```

Interestingly breakding down the missing variables by year confirms what was seen previously. The most reliable variables will be Education Index, Mean Years of Schooling, and Expected Years of Schooling, according to the lower number of missings throughout the years, as well as the variable for population with some secondary education. It also indicates the data is largely lagged as 2018 data seems to be missing while for various variables while previous years have data.

```{r,fig.height=25,fig.width=20}
df<-df_ed_td[c(3:28)] %>% group_by(Year) %>% select(everything()) %>%
     summarise_all(funs(sum(is.na(.)))) 
df$Year<-factor(df$Year)
ggplot(df %>%
     pivot_longer(cols=!c("Year"),names_to="Variable",values_to="Val") ,aes(y=Year,x=Variable,fill=Val))+
  geom_tile()+xlab("Variable")+ylab("Year")+ggtitle("Missing by Year and Variable")+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_fill_continuous(high = "#132B43", low = "#56B1F7")

```
