---
title: "Human.Security"
author: "Michelle Strayer"
---

Below is the code required to compose and organize the Human Security dataset.

Transforming this data largely required reorganizing the dataframe into a tidy
format.  A great number of data values are missing, discussed below.  Converting
the data from its .csv file into a workable R dataframe required removing 
blank columns (which had been left in the Excel files for visual appeal) and 
renaming columns to prime them for the pivot_longer function.  Further, to 
avoid losing data when merging the Human Security variable datasets (Homeless-
ness from natural disasters, Homicide rate, Prison population, Refugees by 
country of origin, Suicide rate for males, and Suicide rate for females), the 
all.x and all.y conditions were specified.  Then, to ensure that all missing
data were coded appropriately, a for loop was introduced to convert all numeric
data columns in the dataset into numeric type.  The result is a tidy dataset of 
all Human Security variables.

```{r}
library(tidyverse)
Homeless.natural.disaster <- read.csv("Homeless.natural.disaster.csv")
Homicide.rate <- read.csv("Homicide.rate.csv")
Prison.popn <- read.csv("Prison.population.csv")
Refugees <- read.csv("Refugees.csv")
Suicide.rate.M <- read.csv("Suicide.rate.M.csv")
Suicide.rate.F <- read.csv("Suicide.rate.F.csv")

names(Homeless.natural.disaster)[2:30] <- c("1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018")

dcol <- 2
for (i in 1:ncol(Homeless.natural.disaster)) {
  Homeless.natural.disaster[,dcol] <- as.numeric(Homeless.natural.disaster[,dcol])
  dcol <- dcol +1
}

Homicide.rate <- subset(Homicide.rate, select = -c(X, X.1, X.2, X.3, X.4, X.5, X.6, X.7, X.8, X.9, X.10))
names(Homicide.rate)[2:13] <- c("1990", "1995", "2000", "2005", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017")

dcol <- 2
for (i in 1:ncol(Homicide.rate)) {
  Homicide.rate[,dcol] <- as.numeric(Homicide.rate[,dcol])
  dcol <- dcol +1
}

Prison.popn <- subset(Prison.popn, select = -c(X, X.1, X.2, X.3, X.4, X.5, X.6, X.7, X.8, X.9))
names(Prison.popn)[2:11] <- c("2003", "2005", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017")

dcol <- 2
for (i in 1:ncol(Prison.popn)) {
  Prison.popn[,dcol] <- as.numeric(Prison.popn[,dcol])
  dcol <- dcol +1
}

Refugees <- subset(Refugees, select = -c(X, X.1, X.2, X.3, X.4, X.5, X.6, X.7, X.8, X.9))
names(Refugees)[2:11] <- c("2003", "2005", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017")

dcol <- 2
for (i in 1:ncol(Refugees)) {
  Refugees[,dcol] <- as.numeric(Refugees[,dcol])
  dcol <- dcol +1
}

Suicide.rate.F <- subset(Suicide.rate.F, select = -c(X, X.1, X.2))
names(Suicide.rate.F)[2:5] <- c("2000", "2010", "2015", "2016")

dcol <- 2
for (i in 1:ncol(Suicide.rate.F)) {
  Suicide.rate.F[,dcol] <- as.numeric(Suicide.rate.F[,dcol])
  dcol <- dcol +1
}

Suicide.rate.M <- subset(Suicide.rate.M, select = -c(X, X.1, X.2))
names(Suicide.rate.M)[2:5] <- c("2000", "2010", "2015", "2016")

dcol <- 2
for (i in 1:ncol(Suicide.rate.M)) {
  Suicide.rate.M[,dcol] <- as.numeric(Suicide.rate.M[,dcol])
  dcol <- dcol +1
}

HND <- pivot_longer(Homeless.natural.disaster, cols = !Country, names_to = "Year", values_to = "Homeless.ND")
HR <- pivot_longer(Homicide.rate, cols = !Country, names_to = "Year", values_to = "Homicide.rate")
PP <- pivot_longer(Prison.popn, cols = !Country, names_to = "Year", values_to = "Prison.popn")
R <- pivot_longer(Refugees, cols = !Country, names_to = "Year", values_to = "Refugees")
SRF <- pivot_longer(Suicide.rate.F, cols = !Country, names_to = "Year", values_to = "Suicide.rate.F")
SRM <- pivot_longer(Suicide.rate.M, cols = !Country, names_to = "Year", values_to = "Suicide.rate.M")

Human.Security <- merge(HND, HR, all.x = T, all.y = T)
Human.Security <- merge(Human.Security, PP, all.x = T, all.y = T)
Human.Security <- merge(Human.Security, R, all.x = T, all.y = T)
Human.Security <- merge(Human.Security, SRF, all.x = T, all.y = T)  
Human.Security <- merge(Human.Security, SRM, all.x = T, all.y = T)
Human.Security <- Human.Security[30:5684,]

```


```{r, echo = FALSE}
write.csv(Human.Security, "Human.Security.csv")
```

Missing Data

A significant proportion of data in the Human.Security dataframe is missing.
Largely, this seems to be the product of uneven variable tracking.  Homelessness
due to natural disasters seems to be tracked in most countries on a yearly
basis between 1990 and 2018.  By contrast, Homicide rate was recorded only in 
1990, 1995, 2000, 2005, and yearly from 2010 to 2017.  As another exmaple, 
suicide rate for both males and females were recorded only in the years 2000, 
2010, 2015, and 2016.

```{r}
colnames(Homeless.natural.disaster)[2:30]
```
```{r}
colnames(Homicide.rate)[2:13]
```
```{r}
colnames(Suicide.rate.F)[2:5]
colnames(Suicide.rate.M)[2:5]
```


Importantly, data also appears to be missing among outlying island countries and
countries in the African continent.  For example, the chart below notes missing 
Homicide data.

A notable chunk of countries (obs. numbers 50-100) seem to be missing data
between 1990 and 2005.  However, the largest amount of missing data seems to 
be concentrated in observations 150 to 192, which includes Mironesia and many of
the countries on and around the African continent (ie. Nigeria, Sudan, Senegal).


```{r}
x <- missing_data.frame(Homicide.rate)
image(x)
```

This trend is not unknown in international data collection.  Lower 
infrastructural development renders data collection difficult, and factors like 
human rights and humanitarian crises, political violence, and political 
instability can hinder data collection processes. 




